Project Rapport\\\textsc{DEVS Ex Machina}

### 2015-04-24 manual/projectrapport.tex
Project rapport

%
\chapter{Summary}
%
\section*{}
The project implements a C++ port of the PyPDEVS\cite{} project, extended with conservative synchronization and a change in system architecture from distributed simulation on several nodes (asymmetric mp) to an symmetric multiprocessing architecture. \\The summarized goal of the project is to provide an implementation of the offered DEVS formalisms with significantly lower execution time by levering the smp architecture.

@@

\chapter{Planning}
\section{Parallel}
\begin {itemize}
\item GVT integration : week 0-1, Ben and Tim
\item Conservative : week 2-3 , Ben and Tim
\end {itemize}
\section{General}
\begin {itemize}
\item Dynamic Structured : week 0-4 Stijn
\item DevStone and examples : week 0-4 Matthijs
\item ......
\end {itemize}
\section{Optimization}
\section{Own Model}

@@

\chapter{Design}
\section{Overview}
We've slightly departed from the PyPDEVS structure, to allow for more encapsulation and decoupling of the core entities. This clear separation of data and responsibilities between classes made the threading part of the project easier (no or minimal shared state).

@@

\section{Network -- Kernel communication}
PyPDEVS uses MPI to exchange messages between kernels, whereas we could use shared memory to communicate between kernels (each running on 1 or more threads). We created a network architecture mimicking the approach used by PyPDEVS, without sacrificing the speed and ordering advantages of shared memory. The Network class consists of a single locked receiving port, and has for each existing kernel an outgoing queue. Receiving a message locks the network for the time it needs to place it on an integer indexed queue (very fast). If a kernel wants to receive messages it can do so if/when it is ready to do so, this only locks if at the same time a message is being received by the network.\\ Several kernels can pull messages lockfree in parallel from the network (since the output queues are not shared). Early benchmarks made a throughput of 5e5 msg/s repeatable, with 4 kernels on modest hardware. The problems of (unintentional) out of order receipt and potential loss of messages are a non-issue in this context, since we use FIFO queues and loss of a message would implicate severe system failure (in which case message loss is the lesser problem).\\
Another advantage of hiding the message exchange behind a shared Network class instance is that, should this be required in the future, the kernel could run on different machines without requiring much redesigning of the system. An RPC implementation would simply have to implement the Network class on both sides.

@@

\subsubsection{Teamwork}
We met at least once every week (barring the exam period), with each meeting having a formal report. Whenever required, teammembers used hangout to solve critical bugs or discuss design/coding issues.
\subsubsection{Code review}
Allthough there was no real formal code review done up to now, we've had several debug sessions reviewing each others code. This means the most critical used sections of the code are allready reasonably well reviewed, albeit not formally.

\chapter{Functional requirements}
\section{DEVS Formalisms}
\section{Single core}
The single core implementations follow exactly the runtime behaviour shown by the PyPDEVS simulator. We've chosen to emulate the Select() function with an integer priority value the user can set. This saves 1 virtual function call per transition and can quite easily be optimized by the compiler since it remains constant.
\subsection{Dynamic Structured}
\subsection{Parallel}
\subsubsection{Optimistic}
Optimistic time synchronization requires an implementation of timewarp, and more extensive locking than would be required for conservative. The GVT algorithm (Mattern \cite{Mattern}) runs on a separate thread from the simulation kernel threads, so care must taken that they don't deadlock, but don't race either.
\subsubsection{Conservative}
Conservative time synchronization is still under development.

\chapter{Obstacles}
\section{Virtualization}
Virtualization proved a significant hurdle by exposing potential faults. Standard compliant code was found to be leaking memory by not calling container member destructors but only on a VM with Linux as Guest and Windows as Host. This was never reprodicible in another setting.
\section{Threading}
G++'s std::string implementation appears to be still Copy On Write (despite 21.4.1 p6 from the C++ standard) \cite{cpp}, triggering an alert by threadsanitizer for a possible race whenever a string was supposedly copied but internally still stored as a reference to the same object. Explicitly invoking the string constructor with the char[] proved to be a safe workaround here. Note that this problem is a known issue \cite{cow}
\section{Toolchains}
To develop we used the required toolchains:\\ g++/clang++ as compiler, Eclipse+CDT as IDE, git for dvcs, jenkins as ci, cmake as build tool supplemented with some shell and python scripting.
\section{Logger}
The recommended logger (g3log), while powerful and fast, proved a dead end since it does not compile on cygwin. 
Stijn therefore wrote his own implementation. %TODO Stijn
\chapter{Performance Analysis}
Whenever we made big design/architecture decisions, we first wrote some prototyping testcode to prove the viability (e.g. Network, Tracing, Logger).
\section{First comparison}
The first performance results were obtained by running PyPDEVS's dynamic structured example against our own simulator, with a termination time of 3.6 E6. It should be noted that the GTEST overhead is not eliminated here yet. The perf \cite{perf} tool was used to obtain low level statistics with as much detail as possible. Unfortunately perf has very limited use on a virtualized OS. The parallel implementation was not yet feature complete, so allthough faster could not be fairly compared to PyPDEVS.
\lstinputlisting[]{pydyn.txt}
The following shows a recording of the C++ implementation of the same example.\\
\lstinputlisting[]{cpdyn.txt}
This result is still warped due to the startup/shutdown time of GTEST, but allready the C++ port is roughly 6x faster, only page faults and migrations are worse (but being worked on).

>>>>

%
\chapter{Summary}
%
\section*{}
The project implements a C++ port of the PyPDEVS project, extended with conservative synchronization and a change in system architecture from distributed simulation on several nodes (asymmetric mp) to a symmetric multiprocessing architecture. \\The summarized goal of the project is to provide an implementation of the offered DEVS formalisms with significantly lower execution time by levering the smp architecture.

@@

\chapter{Planning}
\section{Parallel}
\begin {itemize}
\item GVT integration : week 0-1, Ben and Tim
\item Conservative : week 2-3 , Ben and Tim
\end {itemize}
\section{General}
\begin {itemize}
\item Dynamic Structured : week 1-3 Stijn
\item DevStone and examples : week 0-2 Matthijs and Pieter
\item Serialization : week 0 Pieter
\item Simulation control : week 1-3 Matthijs and Pieter  
\item Tracers : week 0 Stijn
\item Benchmarking PyPDevs Adevs : week 3 Matthijs and Pieter
\end {itemize}
\section{Optimization}
Week 4 , general review.

@@

\chapter{Design}
\section{Overview}
We've slightly departed from the PyPDEVS structure, to allow for more encapsulation and decoupling of the core entities. This clear separation of data and responsibilities between classes made the threading part of the 
project easier (no or minimal shared state).
%TODO include UML overview

@@

\section{Network -- Kernel communication}
PyPDEVS uses MPI to exchange messages between kernels, whereas we could use shared memory to communicate between kernels (each running on 1 or more threads). We created a network architecture mimicking the approach used by PyPDEVS, without sacrificing the speed and ordering advantages of shared memory. The Network class consists of a single locked receiving port, and has for each existing kernel an outgoing queue. Receiving a message locks the network for the time it needs to place it on an integer indexed queue (very fast). If a kernel wants to receive messages it can do so if/when it is ready to do so, this only locks if at the same time a message is being received by the network.\\ Several kernels can pull messages lockfree in parallel from the network (since the output queues are not shared). Early benchmarks made a throughput of 5e5 msg/s repeatable, with 4 kernels on modest hardware. The problems of (unintentional) out of order receipt and potential loss of messages are a non-issue in this context, since we use FIFO queues and loss of a message would implicate severe system failure (in which case message loss is the lesser problem).\\
Another advantage of hiding the message exchange behind a shared Network class instance is that, should this be required in the future, the kernel could run on different machines without requiring much redesigning of the system. An RPC implementation would simply have to implement the Network class on both sides.
\includegraphics[scale=0.2]{network.png}

@@

\subsubsection{Teamwork}
We met at least once every week (barring the exam period), with each meeting having a formal report. Whenever required, teammembers used video conferencing to solve critical bugs or discuss design/coding issues.
\subsubsection{Code review}
Although there was no real formal code review done up to now, we've had several debug sessions reviewing each others code. This means the most critical used sections of the code are allready reasonably well reviewed, albeit not formally.
\subsubsection{Portability}
We chose for Cygwin instead of MinGW because std::thread is not supported on MinGW. 
%TODO Tim gnu++11 and stuff

\chapter{Functional requirements}
\section{DEVS Formalisms}
\section{Single core}
The single core implementations follow exactly the runtime behaviour shown by the PyPDEVS simulator. We've chosen to emulate the Select() function with an integer priority value the user can set. This saves 1 virtual function call per transition and can quite easily be optimized by the compiler since it remains constant.
\subsection{DirectConnect}
%TODO Directconnect : what is rootmodel, only atomics in core.
\subsection{Dynamic Structured}
This is implemented but removing models is not yet fully tested.
%TODO Stijn
\subsection{Parallel}
\subsubsection{Optimistic}
Optimistic time synchronization requires an implementation of timewarp, and more extensive locking than would be required for conservative. The GVT algorithm (Mattern \cite{Mattern}) runs on a separate thread from the simulation kernel threads, so care must taken that they don't deadlock, but don't race either.
\subsubsection{Conservative}
Conservative time synchronization is still under development.

\chapter{Obstacles}
\section{Virtualization}
Virtualization proved a significant hurdle by exposing potential faults. Standard compliant code was found to be leaking memory by not calling container member destructors but only on a VM with Linux as Guest and Windows as Host. This was never reproducible in another setting.
\section{Threading}
G++'s std::string implementation appears to be still Copy On Write (despite 21.4.1 p6 from the C++ standard) \cite{cpp}, triggering an alert by threadsanitizer for a possible race whenever a string was supposedly copied but internally still stored as a reference to the same object. Explicitly invoking the string constructor with the char[] proved to be a safe workaround here. Note that this problem is a known issue \cite{cow}
\section{Toolchains}
To develop we used the required toolchains:\\ g++/clang++ as compiler, Eclipse+CDT as IDE, git for dvcs, jenkins as ci, cmake as build tool supplemented with some shell and python scripting.
\section{Logger}
The recommended logger (g3log), while powerful and fast, proved a dead end since it does not compile on cygwin. 
Stijn therefore wrote his own implementation. %TODO Stijn
\chapter{Performance Analysis}
Whenever we made big design/architecture decisions, we first wrote some prototyping testcode to prove the viability (e.g. Network, Tracing, Logger).
\section{First comparison}
The first performance results were obtained by running PyPDEVS's dynamic structured example against our own simulator, with a termination time of 3.6 E6. It should be noted that the GTEST overhead is not eliminated here yet. The perf \cite{perf} tool was used to obtain low level statistics with as much detail as possible. Unfortunately perf has very limited use on a virtualized OS. The parallel implementation was not yet feature complete, so although faster could not be fairly compared to PyPDEVS.
\lstinputlisting[]{pydyn.txt}
The following shows a recording of the C++ implementation of the same example.\\
\lstinputlisting[]{cpdyn.txt}
This result is still warped due to the startup/shutdown time of GTEST, but allready the C++ port is roughly 6x faster, only page faults and migrations are worse (but being worked on).
